Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-09-16 source-iclr/app_techlemma.tex
proof th1 minor

% !TEX ROOT=./supp-main.tex
To facilitate reading, theorems from the main paper are restated and numbered identically. We first summarize some properties of $L$-smooth and $\mu$-strongly
convex functions~\cite{rockafellar1970convex}. 
\begin{lemma}
	Let $F$ be a convex $L$-smooth function. Then we have the following
	inequalities:

@@

	4. Optimality gap: $F(\vw)-F(\vw^{\ast})\leq$$\langle\nabla F(\vw),\vw-\vw^{\ast}\rangle$.
\end{lemma}
%
\begin{lemma}
	Let $F$ be a $\mu$-strongly convex function. Then 
	\begin{align*}
	F(\vw) & \leq F(\vw')+\langle\nabla F(\vw'),\vw-\vw'\rangle+\frac{1}{2\mu}\|\nabla F(\vw)-\nabla F(\vw')\|^{2}\\
	F(\vw)-F(\vw^{\ast}) & \leq\frac{1}{2\mu}\|\nabla F(\vw)\|^{2}
	\end{align*}
\end{lemma}

\begin{thm}
	Let $\overline{\mathbf{w}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{w}_{T}^{k}$,
	$\nu_{\max}=\max_{k}Np_{k}$, and set decaying learning rates $\alpha_{t}=\frac{1}{4\mu(\gamma+t)}$
	with $\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then
	under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right)
	\end{align*}
	and with partial device participation with at most $K$ sampled devices
	at each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right)
	\end{align*}
\end{thm}

\begin{proof}
	The first step is to observe that the $L$-smoothness of $F$ provides
	the upper bound
	\begin{align*}
	\mathbb{E}(F(\ov{w}_{t}))-F^{\ast} & =\mathbb{E}(F(\ov{w}_{t})-F(\vw^{\ast}))\\
	& \leq\frac{L}{2}\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}
	\end{align*}
	and bound $\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}$.

	Our main step is to prove the bound 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq(1-\mu\alpha_{t})\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+5E^{2}L\alpha_{t}^{3}G^{2}
	\end{align*}

	We have 
	\begin{align*}
	\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & =\|(\ov{w}_{t}-\alpha_{t}\vg_{t})-\vw^{\ast}\|^{2}\\
	& =\|(\ov{w}_{t}-\alpha_{t}\ov{g}_{t}-\vw^{\ast})-\alpha_{t}(\vg_{t}-\ov{g}_{t})\|^{2}\\
	& =A_{1}+A_{2}+A_{3}
	\end{align*}
	where 
	\begin{align*}
	A_{1} & =\|\ov{w}_{t}-\vw^{\ast}-\alpha_{t}\ov{g}_{t}\|^{2}\\
	A_{2} & =2\alpha_{t}\langle\ov{w}_{t}-\vw^{\ast}-\alpha_{t}\ov{g}_{t},\ov{g}_{t}-\vg_{t}\rangle\\
	A_{3} & =\alpha_{t}^{2}\|\vg_{t}-\ov{g}_{t}\|^{2}
	\end{align*}
	By definition of $\vg_{t}$ and $\ov{g}_{t}$ (see \eq{\ref{eq:gradient}}), we have $\mathbb{E}A_{2}=0$.
	For $A_{3}$, we have the follow upper bound:
	\begin{align*}
	\alpha_{t}^{2}\mathbb{E}\|\vg_{t}-\ov{g}_{t}\|^{2} & =\alpha_{t}^{2}\mathbb{E}\|\vg_{t}-\mathbb{E}\vg_{t}\|^{2}=\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}^{2}\|\vg_{t,k}-\mathbb{E}\vg_{t,k}\|^{2}\leq\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}^{2}\sigma_{k}^{2}
	\end{align*}
	again by Jensen's inequality and using the independence of $\vg_{t,k},\vg_{t,k'}$ \cite[Lemma 2]{li2019convergence}.

	Next we bound $A_{1}$: 
	\begin{align*}
	\|\ov{w}_{t}-\vw^{\ast}-\alpha_{t}\ov{g}_{t}\|^{2} & =\|\ov{w}_{t}-\vw^{\ast}\|^{2}+2\langle\ov{w}_{t}-\vw^{\ast},-\alpha_{t}\ov{g}_{t}\rangle+\|\alpha_{t}\ov{g}_{t}\|^{2}
	\end{align*}
	and we will show that the third term $\|\alpha_{t}\ov{g}_{t}\|^{2}$
	can be canceled by an upper bound of the second term. %
	\begin{comment}
	The last term is straightforward to bound by the convexity of $\|\cdot\|^{2}$
	and $L$-smoothness of $F_{k}$,
	\begin{align*}
	\alpha_{t}^{2}\|\ov{g}_{t}\|^{2} & \leq\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}\|\nabla F_{k}(\vw_{t}^{k})\|^{2}\leq2L\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}(F_{k}(\vw_{t}^{k})-F_{k}^{\ast})
	\end{align*}
	or 
	\begin{align*}
	\alpha_{t}^{2}\|\ov{g}_{t}\|^{2} & \leq\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}\|\nabla F_{k}(\vw_{t}^{k})\|^{2}\leq\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}\mathbb{E}\|\nabla F_{k}(\vw_{t}^{k},\mathbf{\xi}_{t}^{k})\|^{2}\leq\alpha_{t}^{2}G^{2}
	\end{align*}
	\end{comment}

	Now 
	\begin{align*}
	& -2\alpha_{t}\langle\ov{w}_{t}-\vw^{\ast},\ov{g}_{t}\rangle\\
	& =-2\alpha_{t}\sum_{k=1}^{N}p_{k}\langle\ov{w}_{t}-\vw^{\ast},\nabla F_{k}(\vw_{t}^{k})\rangle\\
	& =-2\alpha_{t}\sum_{k=1}^{N}p_{k}\langle\ov{w}_{t}-\vw_{t}^{k},\nabla F_{k}(\vw_{t}^{k})\rangle-2\alpha_{t}\sum_{k=1}^{N}p_{k}\langle \vw_{t}^{k}-\vw^{\ast},\nabla F_{k}(\vw_{t}^{k})\rangle\\
	& \leq-2\alpha_{t}\sum_{k=1}^{N}p_{k}\langle\ov{w}_{t}-\vw_{t}^{k},\nabla F_{k}(\vw_{t}^{k})\rangle+2\alpha_{t}\sum_{k=1}^{N}p_{k}(F_{k}(\vw^{\ast})-F_{k}(\vw_{t}^{k}))-\alpha_{t}\mu\sum_{k=1}^{N}p_{k}\|\vw_{t}^{k}-\vw^{\ast}\|^{2}\\
	& \leq2\alpha_{t}\sum_{k=1}^{N}p_{k}\left[F_{k}(\vw_{t}^{k})-F_{k}(\ov{w}_{t})+\frac{L}{2}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}+F_{k}(\vw^{\ast})-F_{k}(\vw_{t}^{k})\right]-\alpha_{t}\mu\|\sum_{k=1}^{N}p_{k}\vw_{t}^{k}-\vw^{\ast}\|^{2}\\
	& =\alpha_{t}L\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}+2\alpha_{t}\sum_{k=1}^{N}p_{k}\left[F_{k}(\vw^{\ast})-F_{k}(\ov{w}_{t})\right]-\alpha_{t}\mu\|\ov{w}_{t}-\vw^{\ast}\|^{2}
	\end{align*}
	For the second term, which is negative, we can ignore it, but this
	yields a suboptimal bound that fails to provide the desired linear
	speedup. Instead, we upper bound it using the following derivation:
	\begin{align*}
	& 2\alpha_{t}\sum_{k=1}^{N}p_{k}\left[F_{k}(\vw^{\ast})-F_{k}(\ov{w}_{t})\right]\\
	& \leq2\alpha_{t}\left[F(\ov{w}_{t+1})-F(\ov{w}_{t})\right]\\
	& \leq2\alpha_{t}\mathbb{E}\langle\nabla F(\ov{w}_{t}),\ov{w}_{t+1}-\ov{w}_{t}\rangle+\alpha_{t}L\mathbb{E}\|\ov{w}_{t+1}-\ov{w}_{t}\|^{2}\\
	& =-2\alpha_{t}^{2}\mathbb{E}\langle\nabla F(\ov{w}_{t}),\vg_{t}\rangle+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}\\
	& =-2\alpha_{t}^{2}\mathbb{E}\langle\nabla F(\ov{w}_{t}),\ov{g}_{t}\rangle+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}\\
	& =-\alpha_{t}^{2}\left[\|\nabla F(\ov{w}_{t})\|^{2}+\|\ov{g}_{t}\|^{2}-\|\nabla F(\ov{w}_{t})-\ov{g}_{t}\|^{2}\right]+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}\\
	& =-\alpha_{t}^{2}\left[\|\nabla F(\ov{w}_{t})\|^{2}+\|\ov{g}_{t}\|^{2}-\|\nabla F(\ov{w}_{t})-\sum_{k}p_{k}\nabla F(\vw_{t}^{k})\|^{2}\right]+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}\\
	& \leq-\alpha_{t}^{2}\left[\|\nabla F(\ov{w}_{t})\|^{2}+\|\ov{g}_{t}\|^{2}-\sum_{k}p_{k}\|\nabla F(\ov{w}_{t})-\nabla F(\vw_{t}^{k})\|^{2}\right]+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}\\
	& \leq-\alpha_{t}^{2}\left[\|\nabla F(\ov{w}_{t})\|^{2}+\|\ov{g}_{t}\|^{2}-L^{2}\sum_{k}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}\right]+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}\\
	& \leq-\alpha_{t}^{2}\|\ov{g}_{t}\|^{2}+\alpha_{t}^{2}L^{2}\sum_{k}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}-\alpha_{t}^{2}\|\nabla F(\ov{w}_{t})\|^{2}
	\end{align*}
	where we have used the smoothness of $F$ twice.

	Note that the term $-\alpha_{t}^{2}\|\ov{g}_{t}\|^{2}$ exactly
	cancels the $\alpha_{t}^{2}\|\ov{g}_{t}\|^{2}$ in the bound
	for $A_{1}$, so that plugging in the bound for $-2\alpha_{t}\langle\ov{w}_{t}-\vw^{\ast},\ov{g}_{t}\rangle$,
	we have so far proved 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+\alpha_{t}L\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}+\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}^{2}\sigma_{k}^{2}\\
	& +\alpha_{t}^{2}L^{2}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}+\alpha_{t}^{3}L\mathbb{E}\|\vg_{t}\|^{2}-\alpha_{t}^{2}\|\nabla F(\ov{w}_{t})\|^{2}
	\end{align*}
	The term $\mathbb{E}\|\vg_{t}\|^{2}\leq G^{2}$ by assumption.

	Now we bound $\mathbb{E}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}$.
	Since communication is done every $E$ steps, for any $t\geq0$, we
	can find a $t_{0}\leq t$ such that $t-t_{0}\leq E-1$ and $\vw_{t_{0}}^{k}=\ov{w}_{t_{0}}$for
	all $k$. Moreover, using $\alpha_{t}$ is non-increasing and $\alpha_{t_{0}}\leq2\alpha{}_{t}$
	for any $t-t_{0}\leq E-1$, we have 
	\begin{align*}
	& \mathbb{E}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}\\
	& =\mathbb{E}\sum_{k=1}^{N}p_{k}\|\vw_{t}^{k}-\ov{w}_{t_{0}}-(\ov{w}_{t}-\ov{w}_{t_{0}})\|^{2}\\
	& \leq\mathbb{E}\sum_{k=1}^{N}p_{k}\|\vw_{t}^{k}-\ov{w}_{t_{0}}\|^{2}\\
	& =\mathbb{E}\sum_{k=1}^{N}p_{k}\|\vw_{t}^{k}-\vw_{t_{0}}^{k}\|^{2}\\
	& =\mathbb{E}\sum_{k=1}^{N}p_{k}\|-\sum_{i=t_{0}}^{t-1}\alpha_{i}\vg_{i,k}\|^{2}\\
	& \leq2\sum_{k=1}^{N}p_{k}\mathbb{E}\sum_{i=t_{0}}^{t-1}E\alpha_{i}^{2}\|\vg_{i,k}\|^{2}\\
	& \leq2\sum_{k=1}^{N}p_{k}E^{2}\alpha_{t_{0}}^{2}G^{2}\\
	& \leq4E^{2}\alpha_{t}^{2}G^{2}
	\end{align*}
	\begin{comment}
	Alternatively, we can also write
	\begin{align*}
	\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2} & =\sum_{k=1}^{N}p_{k}\|\ov{w}_{t-1}-\alpha_{t-1}\vg_{t-1}-\vw_{t-1}^{k}+\alpha_{t-1}\vg_{t-1,k}\|^{2}\\
	& \leq2\sum_{k=1}^{N}p_{k}\left(\|\ov{w}_{t-1}-\vw_{t-1}^{k}\|^{2}+\|\alpha_{t-1}\vg_{t-1}-\alpha_{t-1}\vg_{t-1,k}\|^{2}\right)
	\end{align*}
	and write 
	\begin{align*}
	\sum_{k}p_{k}\|\vg_{t-1,k}-\vg_{t-1}\|^{2}\leq\sum_{k}p_{k}\|\vg_{t-1,k}\|^{2} & =\sum_{k}p_{k}\|\vg_{t-1,k}-\nabla F_{k}(\vw_{t-1}^{k})+\nabla F_{k}(\vw_{t-1}^{k})\|^{2}\\
	& \leq2\sum_{k}p_{k}\|\vg_{t-1,k}-\nabla F_{k}(\vw_{t-1}^{k})\|^{2}+2\sum_{k}p_{k}\|\nabla F_{k}(\vw_{t-1}^{k})\|^{2}\\
	& \leq2\sum_{k}p_{k}\sigma_{k}^{2}+2\sum_{k}p_{k}\|\nabla F_{k}(\vw_{t-1}^{k})-\nabla F_{k}(\ov{w}_{t-1})+\nabla F_{k}(\ov{w}_{t-1})\|^{2}\\
	& \leq2\sum_{k}p_{k}\sigma_{k}^{2}+4\sum_{k}p_{k}\left(\|\nabla F_{k}(\vw_{t-1}^{k})-\nabla F_{k}(\ov{w}_{t-1})\|^{2}+\|\nabla F_{k}(\ov{w}_{t-1})\|^{2}\right)\\
	& \leq2\sigma^{2}+4\sum_{k}p_{k}\left(L^{2}\|\ov{w}_{t-1}-\vw_{t-1}^{k}\|^{2}+\|\nabla F_{k}(\ov{w}_{t-1})\|^{2}\right)
	\end{align*}
	so that 
	\begin{align*}
	\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2} & \leq2(1+4L^{2}\alpha_{t-1}^{2})\sum_{k=1}^{N}p_{k}\|\ov{w}_{t-1}-\vw_{t-1}^{k}\|^{2}+4\sigma^{2}\alpha_{t-1}^{2}+8\alpha_{t-1}^{2}\sum_{k}p_{k}\|\nabla F_{k}(\ov{w}_{t-1})\|^{2}
	\end{align*}
	and taking expectation, we have 
	\begin{align*}
	\mathbb{E}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2} & \leq2(1+4L^{2}\alpha_{t-1}^{2})\mathbb{E}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t-1}-\vw_{t-1}^{k}\|^{2}+4\alpha_{t-1}^{2}(\sigma^{2}+2G^{2})
	\end{align*}
	Using this recusive relation at most $E$ times to arrive at a communication
	round and that $\alpha_{t}L\leq1/8$ for all $t$, we have 
	\begin{align*}
	\mathbb{E}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}\leq E\frac{1}{L^{2}}(\sigma^{2}+2G^{2})
	\end{align*}
	which gives 
	\end{comment}
	{}

	Using the bound on $\mathbb{E}\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2}$,
	we can conclude that, with $\nu_{max}:=N\cdot\max_{k}p_{k}$ and $\nu_{min}:=N\cdot\min_{k}p_{k}$,

	\begin{align*}
	& \mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}^{2}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
	& =\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N^{2}}\sum_{k=1}^{N}(p_{k}N)^{2}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N^{2}}\nu_{max}^{2}\sum_{k=1}^{N}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+6E^{2}L\alpha_{t}^{3}G^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}
	\end{align*}
	where in the last inequality we use $\sigma^{2}=\max_{k}\sigma_{k}^{2}$, and assume $\alpha_{t}$
	satisfies $L\alpha_{t}\leq\frac{1}{8}$. We show next that $\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}=O(\frac{1}{tN}+\frac{E^{2}LG^{2}}{t^{2}})$.

	Let $C\equiv6E^{2}LG^{2}$ and $D\equiv\frac{1}{N}\nu_{max}^{2}\sigma^{2}$.
	Suppose that we have shown $\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}\leq b\cdot(\alpha_{t}D+\alpha_{t}^{2}C)$
	for some constant $b$ and $\alpha_{t}$. Then 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq b(1-\mu\alpha_{t})(\alpha_{t}D+\alpha_{t}^{2}C)+\alpha_{t}^{2}D+\alpha_{t}^{3}C\\
	& =(b(1-\mu\alpha_{t})+\alpha_{t})\alpha_{t}D+(b(1-\mu\alpha_{t})+\alpha_{t})\alpha_{t}^{2}C
	\end{align*}
	and so it remains to choose $\alpha_{t}$ and $b$ such that $(b(1-\mu\alpha_{t})+\alpha_{t})\alpha_{t}\leq b\alpha_{t+1}$
	and $(b(1-\mu\alpha_{t})+\alpha_{t})\alpha_{t}^{2}\leq b\alpha_{t+1}^{2}$.
	Recall that we require $\alpha_{t_{0}}\leq2\alpha_{t}$ for any $t-t_{0}\leq E-1$,
	and $L\alpha_{t}\leq\frac{1}{8}$. If we let $\alpha_{t}=\frac{4}{\mu(t+\gamma)}$
	where $\gamma=\max\{E,32\kappa\}$, then we may check that $\alpha_{t}$
	satisfies both requirements.

	Setting $b=\frac{4}{\mu}$, we have 
	\begin{align*}
	(b(1-\mu\alpha_{t})+\alpha_{t})\alpha_{t} & =\left(b(1-\frac{4}{t+\gamma})+\frac{4}{\mu(t+\gamma)}\right)\frac{4}{\mu(t+\gamma)}\\
	& =\left(b\frac{t+\gamma-4}{t+\gamma}+\frac{4}{\mu(t+\gamma)}\right)\frac{4}{\mu(t+\gamma)}\\
	& =b(\frac{t+\gamma-3}{t+\gamma})\frac{4}{\mu(t+\gamma)}\\
	& \leq b(\frac{t+\gamma-1}{t+\gamma})\frac{4}{\mu(t+\gamma)}\\
	& \leq b\frac{4}{\mu(t+\gamma+1)}=b\alpha_{t+1}
	\end{align*}
	and 
	\begin{align*}
	(b(1-\mu\alpha_{t})+\alpha_{t})\alpha_{t}^{2} & =\left(b(1-\frac{4}{t+\gamma})+\frac{4}{\mu(t+\gamma)}\right)\frac{16}{\mu^{2}(t+\gamma)^{2}}\\
	& =\left(b\frac{t+\gamma-4}{t+\gamma}+\frac{4}{\mu(t+\gamma)}\right)\frac{16}{\mu^{2}(t+\gamma)^{2}}\\
	& =b(\frac{t+\gamma-2}{t+\gamma})\frac{16}{\mu^{2}(t+\gamma)^{2}}\\
	& \leq b\frac{16}{\mu^{2}(t+\gamma+1)^{2}}=b\alpha_{t+1}^{2}
	\end{align*}
	where we have used the facts that 
	\begin{align*}
	\frac{t+\gamma-1}{(t+\gamma)^{2}} & \leq\frac{1}{(t+\gamma+1)}\\
	\frac{t+\gamma-2}{(t+\gamma)^{3}} & \leq\frac{1}{(t+\gamma+1)^{2}}
	\end{align*}
	for $\gamma\geq1$.

	Thus we have shown 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq b\cdot(\alpha_{t+1}D+\alpha_{t+1}^{2}C)
	\end{align*}
	for our choice of $\alpha_{t}$ and $b$. Now to ensure 
	\begin{align*}
	\|\vw_{0}-\vw^{\ast}\|^{2} & \leq b\cdot(\alpha_{0}D+\alpha_{0}^{2}C)\\
	& =b\cdot(\frac{4}{\mu\gamma}D+\frac{16}{\mu^{2}\gamma^{2}}C)
	\end{align*}
	we can simply scale $b$ by $c\|\vw_{0}-\vw^{\ast}\|^{2}$ for a constant
	$c$ large enough and the induction step still holds.

	It follows that 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2} & \leq c\|\vw_{0}-\vw^{\ast}\|^{2}\frac{4}{\mu}(D\alpha_{t}+C\alpha_{t}^{2})
	\end{align*}
	for all $t\geq0$.

	Finally, the $L$-smoothness of $F$ implies 
	\begin{align*}
	& \mathbb{E}(F(\ov{w}_{T}))-F^{\ast}\\
	& =\mathbb{E}(F(\ov{w}_{T})-F(\vw^{\ast}))\\
	& \leq\frac{L}{2}\mathbb{E}\|\ov{w}_{T}-\vw^{\ast}\|^{2}\leq\frac{L}{2}c\|\vw_{0}-\vw^{\ast}\|^{2}\frac{4}{\mu}(D\alpha_{T}+C\alpha_{T}^{2})\\
	& =2c\|\vw_{0}-\vw^{\ast}\|^{2}\kappa(D\alpha_{T}+C\alpha_{T}^{2})\\
	& \leq2c\|\vw_{0}-\vw^{\ast}\|^{2}\kappa\left[\frac{4}{\mu(T+\gamma)}\cdot\frac{1}{N}\nu_{max}^{2}\sigma^{2}+6E^{2}LG^{2}\cdot(\frac{4}{\mu(T+\gamma)})^{2}\right]\\
	& =O(\frac{\kappa}{\mu}\frac{1}{N}\nu_{max}^{2}\sigma^{2}\cdot\frac{1}{T}+\frac{\kappa^{2}}{\mu}E^{2}G^{2}\cdot\frac{1}{T^{2}})
	\end{align*}

	With partial participation, the update at each communication round
	is now given by averages over a subset of sampled devices. When $t+1\notin\mathcal{I}_{E}$,
	$\ov{v}_{t+1}=\ov{w}_{t+1}$, while when $t+1\notin\mathcal{I}_{E}$,
	we have $\mathbb{E}\ov{w}_{t+1}=\ov{v}_{t+1}$ by design
	of the sampling schemes, so that 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & =\mathbb{E}\|\ov{w}_{t+1}-\ov{v}_{t+1}+\ov{v}_{t+1}-\vw^{\ast}\|^{2}\\
	& =\mathbb{E}\|\ov{w}_{t+1}-\ov{v}_{t+1}\|^{2}+\mathbb{E}\|\ov{v}_{t+1}-\vw^{\ast}\|^{2}
	\end{align*}
	As before, $\mathbb{E}\|\ov{v}_{t+1}-\vw^{\ast}\|^{2}\leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+6E^{2}L\alpha_{t}^{3}G^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}$.

	The key is to bound $\mathbb{E}\|\ov{w}_{t+1}-\ov{v}_{t+1}\|^{2}$.
	For sampling scheme I we have 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\ov{v}_{t+1}\|^{2} & =\frac{1}{K}\sum_{k}p_{k}\mathbb{E}\|\vw_{t+1}^{k}-\ov{w}_{t+1}\|^{2}\\
	& \leq\frac{4}{K}\alpha_{t}^{2}E^{2}G^{2}
	\end{align*}
	while for sampling scheme II 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\ov{v}_{t+1}\|^{2} & =\frac{N-K}{N-1}\frac{1}{K}\sum_{k}p_{k}\mathbb{E}\|\vw_{t+1}^{k}-\ov{w}_{t+1}\|^{2}\\
	& \leq\frac{N-K}{N-1}\frac{4}{K}\alpha_{t}^{2}E^{2}G^{2}
	\end{align*}
	The same argument as the full participation case implies 
	\begin{align*}
	\mathbb{E}F(\ov{w}_{T})-F^{\ast}=O(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}

	\begin{comment}
	\begin{proof}
	With partial participation, $2E\sum_{\tau=1}^{E-1}\alpha_{t-\tau}^{2}\sum_{k=1}^{N}p_{k}\left(\|\nabla F_{k}(\ov{w}_{t-\tau},\mathbf{\xi}_{t-\tau}^{k})\|^{2}+l^{2}\|w_{t-\tau}^{k}-\ov{w}_{t-\tau}\|^{2}\right)$
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+5E^{2}L\alpha_{t}^{3}G^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+\frac{1}{K}\sum_{k}p^{k}\|\vw_{t}^{k}-\ov{w}_{t}\|^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})\|\ov{w}_{t}-\vw^{\ast}\|^{2}+5E^{2}L\alpha_{t}^{3}G^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+\frac{4}{K}\alpha_{t}^{2}E^{2}G^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})(1-\mu\alpha_{t-1})\cdots(1-\mu\alpha_{t-E})\|\ov{w}_{t-E}-\vw^{\ast}\|^{2}=O(\frac{1}{t-E}\sigma^{2}+E^{2}LG^{2}\frac{1}{(t-E)^{2}}+\frac{E^{2}G^{2}}{K}\frac{1}{(t-E)^{3/2}})\\
	& +(\alpha_{t}^{3}+(1-\mu\alpha_{t-1})\alpha_{t-1}^{3}+(1-\mu\alpha_{t-1})(1-\mu\alpha_{t-2})\alpha_{t-2}^{3}+\cdots+(1-\mu\alpha_{t-1})\cdots(1-\mu\alpha_{t-E})\alpha_{t-E}^{3}E^{2}LG^{2}\\
	& +(\alpha_{t}^{2}+(1-\mu\alpha_{t-1})\alpha_{t-1}^{2}+(1-\mu\alpha_{t-1})(1-\mu\alpha_{t-2})\alpha_{t-2}^{2}+\cdots+(1-\mu\alpha_{t-1})\cdots(1-\mu\alpha_{t-E})\alpha_{t-E}^{2}\sigma^{2}\\
	& +\frac{4}{K}EG^{2}(\alpha_{t}^{2}+\alpha_{t-1}^{2}+\cdots+\alpha_{t-E}^{2})
	\end{align*}

	so need ot check 
	\begin{align*}
	(1-\mu\alpha_{t})(1-\mu\alpha_{t-1})\cdots(1-\mu\alpha_{t-E+1})b\alpha_{t-E+1}^{\beta}+\alpha_{t}^{2} & \leq\alpha_{t+1}^{\beta}\\
	\end{align*}

>>>>

To facilitate reading, we first summarize some technical lemmas regarding the properties of $L$-smooth and $\mu$-strongly
convex functions~\cite{rockafellar1970convex} which are useful for the proof later on. 
\begin{lemma}
	Let $F$ be a convex $L$-smooth function. Then we have the following
	inequalities:

@@

	4. Optimality gap: $F(\vw)-F(\vw^{\ast})\leq$$\langle\nabla F(\vw),\vw-\vw^{\ast}\rangle$.
\end{lemma}
%
\begin{lemma}
	Let $F$ be a $\mu$-strongly convex function. Then 
	\begin{align*}
	& F(\vw)  \leq F(\vw')+\langle\nabla F(\vw'),\vw-\vw'\rangle+\frac{1}{2\mu}\|\nabla F(\vw)-\nabla F(\vw')\|^{2}\\
	& F(\vw)-F(\vw^{\ast})  \leq\frac{1}{2\mu}\|\nabla F(\vw)\|^{2}
	\end{align*}
\end{lemma}
